Accelerated oblique random survival forests

### 2022-07-14 docs/index.Rmd
results for VI; viz in prog
paper update, slide fix

tar_load(
  names = c(
    bm_pred_model_viz,
    bm_pred_time_viz,
    bm_pred_viz,
    bm_vi_smry,
    data_key
  )
)

@@

```{r bm_pred_viz_ibs, fig.height=9, fig.width=11}
bm_pred_viz$ibs_scaled
```

@@

```{r bm_pred_viz_cstat, fig.height=9, fig.width=11}
bm_pred_viz$cstat
```

@@

```{r bm_pred_model_viz_ibs, fig.height=6.9, fig.width=9}
bm_pred_model_viz$ibs_scaled
```

@@

```{r bm_pred_model_viz_cstat, fig.height=6.9, fig.width=9}
bm_pred_model_viz$cstat
```

@@

```{r bm_pred_time_ridges, fig.height=6.5, fig.width=8}
bm_pred_time_viz
```

@@

fig_data <- bm_vi_smry %>%
  ungroup() %>%
  filter(n_obs == 5000) %>%
  mutate(
    model = recode(
      model, 'randomForestSRC-permutation' = 'rfsrc-permutation'
    ),
    model = fct_reorder(model, .x = overall),
    pred_corr_max = factor(pred_corr_max),
    text = table_glue(
      "{model}<br>\\
          Mean computation time: {as.numeric(time)} seconds<br>\\
          C-statistic by variable class:<br>\\
          Main effect: {main}<br>\\
          Non-linear effect: {nlin}<br>\\
          Combination effect: {cmbn}<br>\\
          Interactions:<br>\\
             w/main effect: {intr_main}<br>\\
             w/non-linear effect: {intr_hidden_nlin}<br>\\
             w/combination effect: {intr_hidden_cmbn}"
    ))

fig <- plot_ly(width = 800, height = 650) %>%
  add_trace(
    type = 'bar',
    x = fig_data$pred_corr_max,
    y = fig_data$overall,
    color = fig_data$model,
    colors = "PuRd",
    text = table_value(fig_data$overall),
    textposition = 'outside',
    hoverinfo = 'text',
    hovertext = fig_data$text
  ) %>%
  plotly::layout(
    title = list(text = 'Results when N = 5000', x = 0.1),
    xaxis = list(title = 'Max correlation among predictors'),
    yaxis = list(title = 'C-statistic', range = c(0,1)),
    legend = list(title = list(text = '<b>VI technique</b>'))
  )

>>>>

tar_load(
  names = c(
    bm_pred_model_viz,
    bm_pred_time_viz,
    bm_pred_viz,
    bm_vi_smry,
    bm_vi_viz,
    data_key
  )
)

@@

```{r bm_pred_viz_ibs, fig.height=9, fig.width=11}
bm_pred_viz$ibs_scaled$fig
```

@@

```{r bm_pred_viz_cstat, fig.height=9, fig.width=11}
bm_pred_viz$cstat$fig
```

@@

```{r bm_pred_model_viz_ibs, fig.height=6.9, fig.width=9}
bm_pred_model_viz$fig$ibs_scaled
```

@@

```{r bm_pred_model_viz_cstat, fig.height=6.9, fig.width=9}
bm_pred_model_viz$fig$cstat
```

@@

```{r bm_pred_time_ridges, fig.height=6.5, fig.width=8}
bm_pred_time_viz$fig
```

@@

fig_data <- bm_vi_smry %>%
  ungroup() %>%
  filter(n_obs == 2500) %>%
  mutate(
    model = recode(
      model, 'randomForestSRC-permutation' = 'rfsrc-permutation'
    ),
    model = fct_reorder(model, .x = overall),
    pred_corr_max = factor(pred_corr_max),
    text = table_glue(
      "{model}<br>\\
          Mean computation time: {as.numeric(time)} seconds<br>\\
          C-statistic by variable class:<br>\\
          Main effect: {main}<br>\\
          Non-linear effect: {nlin}<br>\\
          Combination effect: {cmbn}<br>\\
          Interactions:<br>\\
             w/main effect: {intr_main}<br>\\
             w/non-linear effect: {intr_hidden_nlin}<br>\\
             w/combination effect: {intr_hidden_cmbn}"
    ))

fig <- plot_ly(width = 800, height = 650) %>%
  add_trace(
    type = 'bar',
    x = fig_data$pred_corr_max,
    y = fig_data$overall,
    color = fig_data$model,
    colors = "PuRd",
    text = table_value(fig_data$overall),
    textposition = 'outside',
    hoverinfo = 'text',
    hovertext = fig_data$text
  ) %>%
  plotly::layout(
    title = list(text = 'Results when N = 2500', x = 0.1),
    xaxis = list(title = 'Max correlation among predictors'),
    yaxis = list(title = 'C-statistic', range = c(0,1)),
    legend = list(title = list(text = '<b>VI technique</b>'))
  )

### 2022-07-14 paper/jmlr/main.Rnw
paper update, slide fix

Overall, \texttt{aorsf-negate} had the highest C-statistic in \Sexpr{n_wins_aorsf} out of \Sexpr{n_vi_tasks} VI tasks, with an overall mean of \Sexpr{cstat_mean_aorsf} (Figure \ref{fig:bm_vi_viz}). Compared to the VI technique with the second highest overall C-statistic (\texttt{\Sexpr{cstat_vi_second_place$model}}), \texttt{aorsf-negate}'s mean was \Sexpr{cstat_aorsf_won_by$absolute} points higher, a relative increase of \Sexpr{cstat_aorsf_won_by$percent}\%. Among each of the four relevant variable classes, \texttt{aorsf-negate} had the highest mean C-statistic, with the greatest advantage of using \texttt{aorsf-negate} occurring among non-linear and combination variables and the smallest advantage for interaction variables.

@@

Accurate risk prediction models have the potential to improve healthcare by directing timely interventions to patients who are most likely to benefit. However, prediction models that cannot be interpreted and explained have no place in clinical practice. The current study advances the oblique RSF, an accurate risk prediction model, several steps towards being an accurate \textit{and} interpretable risk prediction model. The improved computational efficiency of the accelerated oblique RSF increases the feasibility of applying model-agnostic methods (\eg, SHAP values) for interpretation. Faster model evaluation and re-fitting also improve diagnosis and resolution of model-based issues (\eg, model calibration deteriorates over time). The introduction of negation VI also advances interpretability. VI is intrinsically linked to model fairness, as it can be used to identify when protected characteristics such as race, religion, and sexuality are inadvertently used (either directly or through correlates of these characteristics) by a prediction model. Since negation VI  engages with the coefficients used in linear combinations of variables, a major component of oblique RSFs, it may be more capable of diagnosing unfairness in oblique RSFs compared to permutation importance and model-agnostic VI techniques.

@@

The current study has several limitations. The accelerated oblique RSF does not account for competing risks, and so our benchmark of prediction accuracy divided tasks where competing risks were present into event-specific tasks. Biased estimation of incidence may occur when competing risks are ignored, and allowing the oblique RSF to account for competing risks is a high priority for future studies. In addition, missing data are not addressed in the accelerated oblique RSF, and users are expected to impute missing values before passing training or testing data into exported functions from the \texttt{aorsf} R package. However, missing data are common and it is standard for ensemble tree methods to handle missing data during the tree growing procedure. Thus, a second item of high priority for future studies is to develop and evaluate strategies to handle missing data while growing an oblique RSF.

>>>>

Overall, \texttt{aorsf-negate} had the highest C-statistic in \Sexpr{n_wins_aorsf} out of \Sexpr{n_vi_tasks} VI tasks, with an overall mean of \Sexpr{cstat_mean_aorsf} (Figure \ref{fig:bm_vi_viz}). Compared to the VI technique with the second highest overall C-statistic (\texttt{\Sexpr{cstat_vi_second_place$model}}), \texttt{aorsf-negate}'s mean was \Sexpr{cstat_aorsf_won_by$absolute} points higher, a relative increase of \Sexpr{cstat_aorsf_won_by$percent}\%. Among the four relevant variable classes, \texttt{aorsf-negate} had the highest mean C-statistic for main effects, combination effects, and non-linear effects, with the greatest advantage of using \texttt{aorsf-negate} occurring among non-linear and combination variables.

The three techniques that used `aorsf` to estimate VI were ranked first (negation), second (ANOVA), and third (permutation) in overall mean C-statistic across all of the simulation scenarios. Computationally, ANOVA VI was the fastest technique with a median average

@@

Accurate risk prediction models have the potential to improve healthcare by directing timely interventions to patients who are most likely to benefit. However, prediction models that cannot be interpreted and explained have no place in clinical practice. The current study advances the oblique RSF, an accurate risk prediction model, towards being accurate, scalable, and interpretable. The improved computational efficiency of the accelerated oblique RSF increases the feasibility of applying oblique RSFs in a wide range of prediction tasks. Faster model evaluation and re-fitting also improve diagnosis and resolution of model-based issues (\eg, model calibration deteriorates over time). The introduction of negation VI also advances interpretability. VI is intrinsically linked to model fairness, as it can be used to identify when protected characteristics such as race, religion, and sexuality are inadvertently used (either directly or through correlates of these characteristics) by a prediction model. Since negation VI  engages with the coefficients used in linear combinations of variables, a major component of oblique RSFs, it may be more capable of diagnosing unfairness in oblique RSFs compared to permutation importance and model-agnostic VI techniques.

@@

The current study has several limitations. The accelerated oblique RSF does not account for competing risks, and so our benchmark of prediction accuracy divided tasks where competing risks were present into event-specific tasks. Biased estimation of incidence may occur when competing risks are ignored, and allowing the oblique RSF to account for competing risks is a high priority for future studies. In addition, missing data are not addressed in the accelerated oblique RSF, and users are expected to impute missing values before passing training or testing data into exported functions from the \texttt{aorsf} R package. However, missing data are common and it is standard for ensemble tree methods to handle missing data during the tree growing procedure. Thus, a second item of high priority for future studies is to develop and evaluate strategies to handle missing data while growing an oblique RSF. Last, \citet{cui2017consistency} found that estimating an inverse-probability weighted hazard function at each non-leaf node of a survival tree allows the RSF to converge asymptotically to the true survival function when some variables contribute both to the risk of the event and the risk of censoring, a scenario that is very likely in the analysis of electronic medical records. The accelerated oblique RSF could incorporate this splitting technique by using Newton Raphson scoring to fit a model for the censoring distribution and then a weighted model could be fit to the failure distribution.
